PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2022-09-28 supplementary.tex
Update on Overleaf.
feat: finished supplementary

% TODO: explain the meaning of L50 and N50

@@

Given a graph composed of many different genomes, PanGraph is capable of marginalizing this graph on a subset of strains with the \verb|marginalize| command. To test this feature we built pangenome graphs on a subset of genomes from the 5 different species we previously analyzed. For each species we took 50 different strains (10 for PM)\footnote{The list of selected accession numbers for each species and pair is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} and built an unique pangenome graph using the \verb|build| command with standard values of the parameters. For each species we then randomly picked 50 different pairs of strains (45 for PM) and extracted the corresponding sub-graphs using the \verb|marginalize| command. We compared each of these graphs to their counterpart generated simply by using the \verb|build| command on the corresponding pair of strains.\\
The comparison between the marginalized (MG) and pairwise (PW) graph consists in splitting each genome using the overlap of the breakpoints given by the two different graphs. Then for each resulting segment we check whether it is shared with the other strain of the pair on either of the graphs. This gives raise to four different possibilities, corresponding to the segment being shared (1) or not (0) on either of the two graphs (MG/PW). We consider four different classes of segments: \textit{agree on sharing} (1-1 or 0-0), \textit{shared on both} (1-1), \textit{private on both} (0-0) and \textit{disagree on sharing} (1-0 or 0-1). For each class we measure the total and average length of the segments belonging to that class, and display the results in Fig.~\ref{fig:marginalize-suppl}.\\
For all species except for PM we find that MG graphs are compatible with their PW counterpart: segments on which the graphs \textit{disagree on sharing} represent only a minor fraction of the genome (much smaller than the fraction of shared or private segments) and their size is compatible with the default block length threshold of PanGraph (100 bp). PM represents and exception, with the great majority of segments being private on each graph and having large size. This expected since for PM the strains considered are highly diverged, above the sensitivity of minimap2 ($\sim 27\%$ core divergence). This causes the majority of blocks not to be merged, and generates a strong order dependence. The size of segments with \textit{disagreement on sharing} is also the largest, some segments having size of 10s of kbp.

\begin{figure}[htb]
    \includegraphics[width=.9\textwidth]{figs_suppl/proj_fig.pdf}
    \caption{{\bf Marginalized graphs are compatible with pairwise graphs}. Comparison of marginalized (MG) and pairwise (PW) built graphs for 5 different species. For each pair of strains we measure the length of segments in which MG and PW graphs agree on whether the segments are shared on both strains (\textit{agree on sharing}). These can either be both shared (\textit{shared on both}) or both private (\textit{private on both}). We also measure the length of segments on which the MG and PW graphs disagree (\textit{disagree on sharing}). For each of these for cases we report the total length as fraction of the average genome (left column) and the average size of each segment (right column). For each category we display the distribution of average results for 50 different pairs (45 for PM).
    }
    \label{fig:marginalize-suppl}
\end{figure}

>>>>

@@

Given a pangenome graph containing many different genomes, \textit{PanGraph} is capable of marginalizing it on a subset of strains with the \verb|marginalize| command. Marginalization removes undesired paths and merges any remaining transitive pancontigs, i.e. pancontigs that are co-linear in all of the remaining paths.
The resulting graph should be equivalent to the one obtained by directly building a graph on the selected subset of strains, with the added advantage that marginalization is a fast operation. This makes it more convenient to build a pangenome graph for the full set of isolates considered, and then marginalizing it multiple times on interesting subsets, rather than building a new graph for each subset.

We used genomes from the dataset described in the previous section to verify that marginalized graphs are compatible with graphs built directly from subset of strains. For each species we selected 50 different isolates (10 for PM) and built pangenome graphs using standard parameter values and the \textit{minimap2} kernel with \textit{asm20} option. For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each pair we then built pairwise pangenome graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf. Fig.~6A, main text).\\
The comparison was performed by considering that each graph partitions a genome into pancontigs. Each pancontig can either be shared with the other member of the pair, or can be private to the isolate (cf. Fig.~6A, main text). For each isolate we consider the two partitions defined by the marginalized and pairwise graph, and generate their intersection. Each block in this intersection can belong to two different categories: either the two original partitions agree or they disagree on whether the block is shared. Moreover, we can separate blocks on which we find agreement in two further categories: blocks that are shared and blocks that are private on both graphs.\\
For each pair of genomes and each type of block we evaluate the fraction of the genome that they cover and their average size. In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered. We consider the four categories described above: blocks on which we find agreement (\textit{agree}) or not (\textit{disagree}), and we separate agreement blocks in \textit{shared} and \textit{private} blocks.\\
In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default block length threhsold of \textit{PanGraph} (100 bp).
As described in the previous section, the divergence of the HP dataset sit on the limit of what the \textit{minimap2} alignment kernel can successfully handle. In this case the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement. In this case the disagreement remains minor and only extends to few percents of the genome size, with disagreement blocks having size only slightly bigger than the 100 bp threshold of PanGraph.
PM represents instead an example of a dataset with divergence much beyond the limit of the alignment kernel ($\dcore \sim 27 \%$). As a consequence, highly diverged homologous sequences are not merged, and the great majority of segments are private and have large size. Results also show a strong order-dependence, with some \textit{disagreement} blocks having size of 10s of kbp.

\begin{figure}[htb]
    \includegraphics[width=.9\textwidth]{figs_suppl/proj_fig.pdf}
    \caption{
        {\bf Marginalized graphs are compatible with pairwise graphs}.
        We compare marginalized graphs with pairwise built graphs for different random pairs of strains from 5 different species, as described in the text. Each graph defines a partition of a genome into segments, that can either be shared in the pair or private to the genome. We intersect these two partitions, and categorize segments based on whether the two partitions \textit{agree} or \textit{disagree} on their sharing. Agreement segments are further split depending on whether the segment is \textit{shared} or \textit{private} on both graphs. The figure displays the distribution of the average size of these segments (right column) and the average fraction of genome that they cover (left column) for the 50 different pairs considered (45 for PM).
        For all species with sequence divergence compatible with the limit of the alignment kernel used (KP, MT, EC), we find very good agreement between marginalized and pairwise graphs, with disagagreement segment only covering a very small fraction of the genome and having size compatible with \textit{PanGraph} threshold pancontig length (100 bp).
    }
    \label{fig:marginalize-suppl}
\end{figure}
